Utilizing Parallel Workers: \\LLVM's Vectorization Plan

### 2024-11-23 paper/paper.tex
paper

LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
Vectorization for outer loops is also in development and can be enabled by setting 
the \texttt{-enable-vplan-native-path} feature flag~\cite{llvmouterloop,llvmouterloopstatus}. 
In the future, the plan will be to merge both of these loop vectorization paths into one. 
Merging the existing SLP Vectorizer with VPlan is also planned.

@@

The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would use VPlans to represent different 
vectorization approaches, abstracted away from the underlying IR. Each VPlan would be able to 
calculate its cost and transform itself to modify the final IR. At first the Legality 
phase (yellow) would check if the scalar code would be legal to vectorize at all and if so initial 
VPlans would be constructed. In the following Planning phase (magenta) these initial VPlans would 
be optimized using various VPlan-to-VPlan transformations. Those transformations would be supervised 
by the VPlan's cost model. Finally, the best VPlan would be chosen and materialized/executed, 
modifying the underlying IR. The best VPlan might also be to not vectorize at all (VF $=1$).

@@

\subsection{State of VPlan (LLVM 19.1.4)}

In its current state VPlan is deeply integrated into LV.
When auto-vectorizing, LV starts by checking all the different loops in the IR.
If it finds one, it will try to vectorize it\\(\texttt{LoopVectorizePass::processLoop}).
Several factors might hinder vectorization for a given loop: 
(1) If the loop was already vectorized before, it won't be vectorized again. 
(2) The Legality phase checks if an how it is legal to vectorize the scalar
loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost
of runtime checks for e.g. pointer aliasing can be too high.

The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
LV cost model made referencing the scalar IR\\(\texttt{LoopVectorizationPlanner::plan}). Then the
optimal Vectorization Factor (VF) and Interleave Count (IC) are picked. 
The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
profitable to vectorize the loop, however, interleaving/unrolling can still end up modifying the
underlying IR.

The best VPlan is then executed\\(\texttt{LoopVectorizationPlanner::executePlan}). In case of
vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.

\subsubsection{VPlan Structure}
To better understand the
underlying structure of a VPlan, modelling a certain vectorization approach, we will use the
simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.

Figure~\ref{fig:inner-loop-scalar-ir} shows the IR of the main loop basic blocks \texttt{for.body},
\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg} shows the equivalent
hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
scalar loop.

%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
%       double column figure.
\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir.png}
  \caption{The scalar IR for the main loop part of the conditional loop depicted in
  Figure~\ref{fig:inner-loop-vec}. Note that \texttt{getelementptr inbounds}
  was abbreviated to \texttt{getelmptr inbs}.}
  \label{fig:inner-loop-scalar-ir}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-2.png}
  \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
  depicted in Figure~\ref{fig:inner-loop-vec}. This VPlan has a VF of 4. Note that some keywords
  are abbreviated: \texttt{arrayidx} $\equiv$ \texttt{arridx}; 
  \texttt{getelementptr inbounds} $\equiv$ \texttt{getelmptr inbs}}
  \label{fig:inner-loop-vplan-hcfg}
\end{figure}

@@

\subsubsection{Outer Loop Vectorization}
Development around outer loop vectorization has slowed down a bit as focus shifted more onto VPlan
internals and integration into LV. The inner loop path and outer loop path are currently still
seperated by the experimental \texttt{-enable-vplan-native-path} flag. Further developing the 
outer loop path and eventually merging it with the innermost loop path is planned.

@@

Outer loop vectorization support via VPlan is still experimental, seperated from the inner loop
LV integration, and runs on a divergent experimental path (\texttt{-enable-vplan-native-path}).
In the near future development on outer loop vectorization will continue and merging it with LV's
VPlan path is planned.

@@

\section{Appendix: Additional Figures}

>>>>

LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
Vectorization for outer loops is also in development and can be enabled by setting 
the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} feature 
flag~\cite{llvmouterloop,llvmouterloopstatus}. 
In the future, the plan will be to merge both of these loop vectorization paths into one. 
Merging the existing SLP Vectorizer with VPlan is also planned.

@@

The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would use VPlans to represent different 
vectorization approaches, abstracted away from the underlying IR. Each VPlan would be able to 
calculate its cost and transform itself to modify the final IR. At first the Legality 
phase (yellow) would check if the scalar code would be legal to vectorize at all and if so initial 
VPlans would be constructed. In the following Planning phase (magenta) these initial VPlans would 
be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
Those transformations would be supervised 
by the VPlan's cost model. Finally, the best VPlan would be chosen and materialized/executed, 
modifying the underlying IR. The best VPlan might also be to not vectorize at all (VF $=1$).

@@

\subsection{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em

In its current state VPlan is deeply integrated into LV.
When auto-vectorizing, LV starts by checking all the different loops in the IR.
If it finds one, it will try to vectorize it (\texttt{LoopVectorizePass::\allowbreak processLoop}).
Several factors might hinder vectorization for a given loop: 
(1) If the loop was already vectorized before, it won't be vectorized again. 
(2) The Legality phase checks if an how it is legal to vectorize the scalar
loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost
of runtime checks for e.g. pointer aliasing can be too high.

The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
LV cost model made referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}).
After initial VPlan construction various VPlan-to-VPlan transformations are applied to optimize
it (\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).

Then the optimal Vectorization Factor (VF) is calculated based on the set of VPlans and their
representive cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), 
which calculation still involves the legacy LV cost model. 
The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
profitable to vectorize the loop, however, interleaving/unrolling can still end up modifying the
underlying IR.
Afterwards, the Interleave Count (IC) is determined 
(\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).

The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In case of
vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.

\subsubsection{VPlan Structure}
To better understand the structure of a VPlan, we will use the
simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.

Figure~\ref{fig:inner-loop-scalar-ir} shows the IR of the main loop basic blocks \texttt{for.body},
\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
scalar loop. Other basic blocks like the scalar epilogue loop and it's predication check (is the final
vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
in Appendix~\ref{appendix-figures} for the fully modeled HCFG.

The hierarchical CFG of a VPlan helps abstract it away from the underlying IR. The nodes of this CFG
are either \texttt{VPBasicBlocks}, containing a sequence of one or more \texttt{VPRecipes}, or
\texttt{VPRegionBlocks}, modelling a single-entry-single-exit subgraph.
In our example the \texttt{vector.body} is a basic block and \texttt{vector loop} is a region block.

\texttt{VPRegionBlocks} can indicate that their contents are to be replicated a certain number of times,
to model more complex control flow for different VF, UF combinations more efficiently in a single VPlan. 
In our example \texttt{vector loop} is not to be replicated (\texttt{<x1>}).

Inside our two \texttt{VPBasicBlocks} we have a list of recipes (lines of instructions). A recipe
models one or more instructions in the executed IR and can be based on one or more instructions.
from the input IR (ingredients). Note that a recipe might also model a instruction with no reference to any input
IR (in our example this would be the inversion of our vector mask). There are different types of 
recipes. \texttt{VPReplicateRecipe} will either \texttt{CLONE} exactly one instruction from the input IR
into exactly one instruction in the output IR, or \texttt{REPLICATE} an instruction a certain number
of times into output IR (typically found in replicator \texttt{VPRegionBlocks}). Different 
\texttt{VPWidenRecipes} exist to transform/\texttt{WIDEN} any input IR instruction into a vectorized 
version of it. The last remaining recipe type in our example are \texttt{VPInstructions}. They
\texttt{EMIT} new instructions into the output IR that have no representative in the input IR (no
ingredients).

Furthermore, VPlan recipes can also model definitions and usages of different values. In our example,
the comparison mask \texttt{vp<\%6>} is a \texttt{VPValue} that is used by three different 
\texttt{VPUsers}.

%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
%       double column figure.
\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir-color.png}
  \caption{The scalar IR for the main loop part of the conditional loop depicted in
  Figure~\ref{fig:inner-loop-vec}. Note that \texttt{getelementptr inbounds}
  was abbreviated to \texttt{getelmptr inbs}.}
  \label{fig:inner-loop-scalar-ir}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
  depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

@@

\subsubsection{Outer Loop Vectorization}
Development around outer loop vectorization has slowed down a bit as focus shifted more onto VPlan
internals and integration into LV. The inner loop path and outer loop path are currently still
seperated by the experimental \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
flag. Further developing the 
outer loop path and eventually merging it with the innermost loop path is planned.

@@

Outer loop vectorization support via VPlan is still experimental, seperated from the inner loop
LV integration, and runs on a divergent experimental path (\texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path}).
In the near future development on outer loop vectorization will continue and merging it with LV's
VPlan path is planned.

@@

\section{Appendix: Additional Figures}\label{appendix-figures}
